The non-independence of nations and why it matters

### 2023-04-04 manuscript.Rmd
second resubmission

keywords          : "spatial non-independence; cultural non-independence; cross-national analyses; simulations; replications; reanalyses"
wordcount         : "4892 words excl. Methods (Methods: 1877 words)"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \raggedbottom
  - \usepackage{setspace}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1.5}}
  - \captionsetup[figure]{font={stretch=1.5}}
appendix          : "appendix.Rmd"
csl               : nature.csl
urlcolor          : blue
---

@@

There are other approaches to controlling for non-independence that we did not include in our simulation. For example, conditional autoregressive models[@Lichstein2002] and generalised additive models[@Hastie2017] are approaches that can be applied in both frequentist and Bayesian frameworks. There are also alternative ways to operationalise cultural distances between nations, including metrics like cultural fixation indices (e.g. cultural $F_{ST}$[@Muthukrishna2020;@White2021]), covariance based on genetic distances[@Spolaore2009], and phylogenetic distances between religious traditions. These metrics allow for cultural ancestry effects beyond ethnolinguistic genealogies, though some may produce less interpretable results than language ancestry. For example, unlike language, genes are strongly correlated with geography[@Handley2007;@Novembre2008] and their global patternings are consistent with different underlying processes, such as deep cultural inheritance and more recent gene flow between populations. Future work should explore the interpretability of these methods and determine whether they are sufficient to reduce false positive rates in spatially and culturally non-independent data. In addition to spatial proximity and shared cultural ancestry, we did not simulate other sources of non-independence that potentially exist in real cross-national datasets, such as modern connections between nations due to flows of people and information (e.g. flight networks, social media networks) and shared histories of colonialism and capitalist modes of production that have shaped today's global landscape[@Trouillot2016;@Wolf2010]. Additional controls will be required to ensure that these sources of non-independence do not confound cross-national inferences.

@@

Until such changes are implemented and sufficient controls for non-independence are the norm, existing cross-national correlations should be interpreted with caution. In our reanalyses, we found that four out of six cross-national correlations with economic development variables had 95% credible intervals that included zero when controlling for spatial non-independence. Three of these analyses were tests of "persistence" hypotheses, studying the effects of historical and environmental conditions --- settler mortality[@Beck2003], wheat-sugar suitability[@Easterly2007], and natural disaster frequency[@Skidmore2002] --- on modern developmental outcomes. These results are in line with a recent paper that has called into question various studies of this ilk[@Kelly2020] and contradict claims that the concern over spatial non-independence has been overblown in these studies[@Voth2021]. We also found that two out of six cross-national correlations with cultural values variables had 95% credible intervals that included zero when controlling for cultural phylogenetic non-independence. Overall, these reanalyses show that the problem of statistical non-independence applies to a wider range of national-level variables than those identified by previous work, such as parasite stress and democratic outcomes[@Bromham2018], raising the worrying possibility that the current cross-national literature is populated with spurious relationships. Future work should expand our set of reanalyses to determine the extent of this problem in the literature.

@@

>>>>

keywords          : "spatial non-independence; cultural non-independence; cross-national analyses; simulations; replications; reanalyses"
wordcount         : "4892 words excl. Methods (Methods: 1877 words)"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \raggedbottom
  - \usepackage{setspace}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1.5}}
  - \captionsetup[figure]{font={stretch=1.5}}
  - \nolinenumbers
appendix          : "appendix.Rmd"
csl               : nature.csl
urlcolor          : blue
---

\linenumbers

@@

There are other approaches to controlling for non-independence that we did not include in our simulation. For example, conditional autoregressive models[@Lichstein2002] and generalised additive models[@Hastie2017] are approaches that can be applied in both frequentist and Bayesian frameworks. There are also alternative ways to operationalise cultural distances between nations beyond linguistic distances, including metrics like cultural fixation indices (e.g. cultural $F_{ST}$[@Muthukrishna2020;@White2021]), covariance based on genetic distances[@Spolaore2009], and phylogenetic distances between religious traditions. We see merit in each of these approaches and the use of one over another will depend on the specific question at hand. We decided to focus on linguistic distances in this study since language is a tangible socially-learned trait that has previously been used to successfully track the effects of deep cultural ancestry on modern national outcomes[@Kyritsis2022;@Matthews2016;@Sookias2018]. Future work should explore whether other approaches are sufficient to reduce false positive rates in spatially and culturally non-independent data. In addition to spatial proximity and shared cultural ancestry, we did not simulate other sources of non-independence that potentially exist in real cross-national datasets, such as modern connections between nations due to flows of people and information (e.g. flight networks, social media networks) and shared histories of colonialism and capitalist modes of production that have shaped today's global landscape[@Trouillot2016;@Wolf2010]. Additional controls will be required to ensure that these sources of non-independence do not confound cross-national inferences.

@@

Until such changes are implemented and sufficient controls for non-independence are the norm, existing cross-national correlations should be interpreted with caution. In our reanalyses, we found that half of the cross-national correlations had 95% credible intervals that included zero when controlling for spatial and/or cultural phylogenetic non-independence. While these results are in line with previous reanalyses[@Kelly2020], we note that we are unable to outright reject the claims from these studies, since we only reanalysed the first bivariate regression specifications presented in the papers[@Voth2021]. More detailed sets of reanalyses would be required to comprehensively challenge the claims from these specific papers. Nevertheless, these reanalyses do show, more broadly, that the problem of statistical non-independence applies to a wider range of national-level variables than those identified by previous work, such as parasite stress and democratic outcomes[@Bromham2018]. Moreover, given our finding that most studies in the current cross-national literature do not deal with non-independence at all, our reanalyses raise the worrying possibility that this literature is populated with spurious relationships. Future work should expand our set of reanalyses to determine the extent of this problem in the literature.

@@

\newpage
\nolinenumbers

### 2023-04-04 manuscript.tex
second resubmission

\linenumbers
\usepackage{csquotes}
\raggedbottom
\usepackage{setspace}
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1.5}}
\captionsetup[figure]{font={stretch=1.5}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The non-independence of nations and why it matters},
  pdfauthor={Scott Claessens1, Thanos Kyritsis1, \& Quentin D. Atkinson1,2},
  pdflang={en-EN},
  pdfkeywords={spatial non-independence; cultural non-independence; cross-national analyses; simulations; replications; reanalyses},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

@@

@@

There are other approaches to controlling for non-independence that we did not include in our simulation. For example, conditional autoregressive models\textsuperscript{31} and generalised additive models\textsuperscript{65} are approaches that can be applied in both frequentist and Bayesian frameworks. There are also alternative ways to operationalise cultural distances between nations, including metrics like cultural fixation indices (e.g.~cultural \(F_{ST}\)\textsuperscript{6,66}), covariance based on genetic distances\textsuperscript{67}, and phylogenetic distances between religious traditions. These metrics allow for cultural ancestry effects beyond ethnolinguistic genealogies, though some may produce less interpretable results than language ancestry. For example, unlike language, genes are strongly correlated with geography\textsuperscript{68,69} and their global patternings are consistent with different underlying processes, such as deep cultural inheritance and more recent gene flow between populations. Future work should explore the interpretability of these methods and determine whether they are sufficient to reduce false positive rates in spatially and culturally non-independent data. In addition to spatial proximity and shared cultural ancestry, we did not simulate other sources of non-independence that potentially exist in real cross-national datasets, such as modern connections between nations due to flows of people and information (e.g.~flight networks, social media networks) and shared histories of colonialism and capitalist modes of production that have shaped today's global landscape\textsuperscript{70,71}. Additional controls will be required to ensure that these sources of non-independence do not confound cross-national inferences.

@@

Until such changes are implemented and sufficient controls for non-independence are the norm, existing cross-national correlations should be interpreted with caution. In our reanalyses, we found that four out of six cross-national correlations with economic development variables had 95\% credible intervals that included zero when controlling for spatial non-independence. Three of these analyses were tests of ``persistence'' hypotheses, studying the effects of historical and environmental conditions --- settler mortality\textsuperscript{53}, wheat-sugar suitability\textsuperscript{57}, and natural disaster frequency\textsuperscript{58} --- on modern developmental outcomes. These results are in line with a recent paper that has called into question various studies of this ilk\textsuperscript{40} and contradict claims that the concern over spatial non-independence has been overblown in these studies\textsuperscript{63}. We also found that two out of six cross-national correlations with cultural values variables had 95\% credible intervals that included zero when controlling for cultural phylogenetic non-independence. Overall, these reanalyses show that the problem of statistical non-independence applies to a wider range of national-level variables than those identified by previous work, such as parasite stress and democratic outcomes\textsuperscript{41}, raising the worrying possibility that the current cross-national literature is populated with spurious relationships. Future work should expand our set of reanalyses to determine the extent of this problem in the literature.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Handley2007}{}}%
\CSLLeftMargin{68. }%
\CSLRightInline{Handley, L. J. L., Manica, A., Goudet, J. \& Balloux, F. Going the distance: Human population genetics in a clinal world. \emph{TRENDS in Genetics} \textbf{23}, 432--439 (2007).}

\leavevmode\vadjust pre{\hypertarget{ref-Novembre2008}{}}%
\CSLLeftMargin{69. }%
\CSLRightInline{Novembre, J. \emph{et al.} Genes mirror geography within europe. \emph{Nature} \textbf{456}, 98--101 (2008).}

@@

\renewcommand{\appendixname}{\bf{Supplementary Material}}
\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}
\renewcommand{\thefigure}{S\arabic{figure}} \setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}} \setcounter{table}{0}
\renewcommand{\theequation}{S\arabic{table}} \setcounter{equation}{0}

>>>>

\linenumbers
\usepackage{csquotes}
\raggedbottom
\usepackage{setspace}
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1.5}}
\captionsetup[figure]{font={stretch=1.5}}
\nolinenumbers
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The non-independence of nations and why it matters},
  pdfauthor={Scott Claessens1, Thanos Kyritsis1, \& Quentin D. Atkinson1,2},
  pdflang={en-EN},
  pdfkeywords={spatial non-independence; cultural non-independence; cross-national analyses; simulations; replications; reanalyses},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

@@

\linenumbers

@@

There are other approaches to controlling for non-independence that we did not include in our simulation. For example, conditional autoregressive models\textsuperscript{31} and generalised additive models\textsuperscript{65} are approaches that can be applied in both frequentist and Bayesian frameworks. There are also alternative ways to operationalise cultural distances between nations beyond linguistic distances, including metrics like cultural fixation indices (e.g.~cultural \(F_{ST}\)\textsuperscript{6,66}), covariance based on genetic distances\textsuperscript{67}, and phylogenetic distances between religious traditions. We see merit in each of these approaches and the use of one over another will depend on the specific question at hand. We decided to focus on linguistic distances in this study since language is a tangible socially-learned trait that has previously been used to successfully track the effects of deep cultural ancestry on modern national outcomes\textsuperscript{62,68,69}. Future work should explore whether other approaches are sufficient to reduce false positive rates in spatially and culturally non-independent data. In addition to spatial proximity and shared cultural ancestry, we did not simulate other sources of non-independence that potentially exist in real cross-national datasets, such as modern connections between nations due to flows of people and information (e.g.~flight networks, social media networks) and shared histories of colonialism and capitalist modes of production that have shaped today's global landscape\textsuperscript{70,71}. Additional controls will be required to ensure that these sources of non-independence do not confound cross-national inferences.

@@

Until such changes are implemented and sufficient controls for non-independence are the norm, existing cross-national correlations should be interpreted with caution. In our reanalyses, we found that half of the cross-national correlations had 95\% credible intervals that included zero when controlling for spatial and/or cultural phylogenetic non-independence. While these results are in line with previous reanalyses\textsuperscript{40}, we note that we are unable to outright reject the claims from these studies, since we only reanalysed the first bivariate regression specifications presented in the papers\textsuperscript{63}. More detailed sets of reanalyses would be required to comprehensively challenge the claims from these specific papers. Nevertheless, these reanalyses do show, more broadly, that the problem of statistical non-independence applies to a wider range of national-level variables than those identified by previous work, such as parasite stress and democratic outcomes\textsuperscript{41}. Moreover, given our finding that most studies in the current cross-national literature do not deal with non-independence at all, our reanalyses raise the worrying possibility that this literature is populated with spurious relationships. Future work should expand our set of reanalyses to determine the extent of this problem in the literature.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Kyritsis2022}{}}%
\CSLLeftMargin{68. }%
\CSLRightInline{Kyritsis, T., Matthews, L. J., Welch, D. \& Atkinson, Q. D. \href{https://doi.org/10.1017/ehs.2022.40}{Shared cultural ancestry predicts the global diffusion of democracy}. \emph{Evolutionary Human Sciences} \textbf{4}, e42 (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-Matthews2016}{}}%
\CSLLeftMargin{69. }%
\CSLRightInline{Matthews, L. J., Passmore, S., Richard, P. M., Gray, R. D. \& Atkinson, Q. D. \href{https://doi.org/10.1371/journal.pone.0152979}{Shared cultural history as a predictor of political and economic changes among nation states}. \emph{PLOS ONE} \textbf{11}, 1--18 (2016).}

@@

\renewcommand{\appendixname}{\bf{Supplementary Material}}
\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}
\renewcommand{\thefigure}{S\arabic{figure}} \setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}} \setcounter{table}{0}
\renewcommand{\theequation}{S\arabic{table}} \setcounter{equation}{0}
\pagenumbering{gobble}
